\bf Promoting analysis reproducibility with accessibility: An example in evolutionary biology

### 2022-02-04 jsdse/Manuscript.tex
adding references

To elaborate on our thesis, we use an example in evolutionary biology that focuses
on understanding the shared ancestry of organisms through phylogenetic trees, which provide the
basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
As such, improving reproducibility and availability
in phylogenetic research is relevant for many aspects of biological research.
The Open Tree of Life project (OpenTree) has developed a platform that facilitates
availability of results from phylogenetic research, by standardizing and storing
phylogenetic data with the goal of synthesizing a single phylogenetic tree encompassing
all life (\cite{opentreeoflife2019synth}).
All data in OpenTree is open access and available programmatically through its many
Application Programming Interface (API) services (\cite{opentreeAPIs}).
Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries
(\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services
to make them available to a wider programming audience.
The R and Python OpenTree wrappers have been utilized by computer-literate individuals,
to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
knowledge for biological research (\cite{sanchez2019datelife}) and education
(\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).

The OpenTree project demonstrates that efforts to increase availability for reproducibility
in phylogenetics have also increased baseline computational knowledge and skills
required in the field.
Baseline computational skill requirements are likely increasing across all natural sciences.
Computing is not traditionally a core skill taught to biologists and naturalists.
However, many students are now being trained in R as a statistics and data analysis environment.
In order for reproducible computational tools to be adopted for research, they
 need to be much more accessible for present and future researchers.
Data and code availability is a core requirement for reproducibility research
 (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
However, the utility of data resources is limited by the technical challenges of accessing the data.
In order to motivate reproducible research, that gap needs to be bridged.
In this work we focus on improving accessibility of code examples and documentation.
In particular, we identify the necessity for documentation that is written down
 using language that is common to the target audience to facilitate examination,
  application, and adoption of code by the wider audience.

@@

Good primary documentation for code describes general usage of individual functions,
 the components and variables a function can take, and it should be accompanied with
  function usage examples on how to apply it.
As opposed to code, primary documentation is written in natural language (i.e.,
 any known human language, e.g., English, Spanish, Chinese) and usually makes use
  of highly specialized computational jargon (computationally specific concepts,
   words, and phrases) as well as formal language, which often slows down or even
    obstructs examination, application, and adoption of code by external individuals.
Because primary documentation is considered a professional document, acceptance
 of the research by the scientific community could be reduced if a more informal
  language is used.
Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
 additional cases of individual function usage, and showcase function associations
  that work for specific analysis workflows in more detail.
   While secondary documentation has become more common practice,
    it is still often generated using highly specialized language.

@@

\subsubsection*{a. Demonstrate integration of function usage with motivating examples}

@@

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.
Many finely honed tutorials do not trigger errors, which precludes helping students
to develop the tools to understand and address errors when they do encounter them,
as they inevitably will.
On our tutorial, we focus on explaining the meaning and downstream coof warnings and errors, and
 showcase ways to detect them before they are triggered (i.e., before using an input
  that would elicit a warning or error). This has two pedagogical benefits:
1) it provides users/students with the means to troubleshoot their own warnings and erros, and
2) it allows users/students to understand with more depth what the function is doing.

>>>>

To elaborate on our thesis, we use an example in evolutionary biology that focuses
on understanding the shared ancestry of organisms through phylogenetic trees, which provide the
basis to study and understand biological processes in an evolutionary context (\cite{dobzhansky1973nothing}).
As such, improving reproducibility rates
in phylogenetic research is relevant for many aspects of biological research.
The Open Tree of Life project (OpenTree) has developed a computational platform that provides
availability of results from phylogenetic research, by creating a standardized database of
phylogenetic data that is used to synthesize a single phylogenetic tree encompassing
all life (\cite{opentreeoflife2019synth}).
The OpenTree database Phylesystem (\cite{}) is open access and available programmatically through its many
Application Programming Interface (API) services (\cite{opentreeAPIs}).
Additionally, R packages (\cite{michonneau2016rotl}) and Python libraries
(\cite{mctavish2021opentree}) have been developed as wrappers for OpenTree API services
to make them available to a wider programming audience.
The R and Python OpenTree wrappers have been utilized by computer-literate individuals,
to seamlessly establish reproducible workflows to use and reuse expert phylogenetic
knowledge for biological research (\cite{sanchez2019datelife}) and education
(\cite{nguyen2020phylotastic, phylotasticedtools, galacticedtools}).

The OpenTree project demonstrates that efforts to increase availability for reproducibility
in phylogenetics have also increased baseline computational knowledge and skills
required in the field.
"Computers are playing an increasingly important role in many scientific disciplines" (\cite{piccolo2016tools}).
Baseline computational skill requirements are increasing across all natural sciences.
Yet, computing is not traditionally a core skill taught to biologists and naturalists
(\cite{sayres2018bioinformatics, wright2019the, williams2019barriers}).
Still, many students are now being trained in R programming language as a statistics and data analysis environment.
% "Thus efforts to increase scientific reproducibility should consider the ubiquity of computers in research."
In order for reproducible computational tools to be adopted for research, they
 need to be much more accessible for present and future researchers.
Data and code availability is a core requirement for reproducibility research
 (\cite{peng2011reproducible, sandve2013ten, powers2019open}).
However, the utility of data resources is limited by the technical challenges of accessing the data.
In order to motivate reproducible research, that gap needs to be bridged.
In this work we focus on improving accessibility of code examples and documentation.
In particular, we identify the necessity for documentation that is written down
 using language that is common to the target audience to facilitate examination,
  application, and adoption of code by the wider audience.

@@

Good primary documentation for code describes general usage of individual functions,
 the components and variables a function can take, and it should be accompanied with
  function usage examples on how to apply it.
As opposed to code, primary documentation is written in natural language (i.e.,
 any known human language, e.g., English, Spanish, Chinese) and usually makes use
  of highly specialized computational jargon (computationally specific concepts,
   words, and phrases) as well as formal language, which often slows down or even
    obstructs examination, application, and adoption of code by external individuals.
Because primary documentation is considered a professional document, acceptance
 of the research by the scientific community could diminish if a more informal
  language is used.
Vignettes and tutorials work as secondary pieces of documentation, that help to demonstrate
 additional cases of individual function usage, and showcase function associations
  that work for specific analysis workflows in more detail.
As secondary documentation has become more common practice, we identify and propose
a series of principles to support improved accessibility for secondary documentation.

@@

\subsubsection*{a. Literate programming: Demonstrate code usage with integrative examples}

%% By definition, all languages are abstract. We learn them because we grow up with them and they have a direct application for our lives.
%% Programming languages are no different, however, we start learning those later in life and we do not use them as regularly, making it harder for us to learn them and use them.
Programming languages are abstract and usually hard to remember without
an obvious link to an application (\cite{knuth1984literate}).
A story or narrative that links pieces of code together can greatly support users/students
to remember the code and its usage.
The "literate programming" paradigm (\cite(knuth1984literate))
integrates narratives with computer code in the same document, to support users
in following the code usage, improving reproducibility of scientific analyses (\cite(piccolo2016tools)).

@@

We identify ways to evaluate inputs to know if they will produce an error, and design
alternatives on what to do when faced with an error or warning, and demonstrate
these alternatives.
One of the most essential skills in programming is interpreting and moving forward
from errors.
Many finely honed tutorials do not trigger errors, which precludes helping students
to develop the tools to understand and address errors when they do encounter them,
as they inevitably will.
On our tutorial, we focus on explaining the meaning and downstream of warnings and errors, and
 showcase ways to detect them before they are triggered (i.e., before using an input
  that would elicit a warning or error). This has two pedagogical benefits:
1) it provides users/students with the means to troubleshoot their own warnings and erros, and
2) it allows users/students to understand with more depth what the function is doing.
